---
title: "Naomi model workflow"
author:
- name: Adam Howes
output:
  html_document:
    toc: yes
    toc_float: true
    toc_collapsed: true
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
    **Background** A vignette for the Naomi model is uploaded to the package website.
    
    **Task** We run through the vignette manually to (1) develop a better understanding of the model, (2) get experience working with the Malawi data, which we will likely use for our case-study as it is publicly available.
---

```{r}
knitr::opts_chunk$set(error = TRUE)

library(naomi)
library(tidyverse)
library(sf)
```

# Naomi vignette

Following vignette [here](https://mrc-ide.github.io/naomi/articles/model-workflow.html).

## Load data inputs

Area hierarchy and boundaries, and plot to check we've got the right thing:

```{r}
area_merged <- read_sf(system.file("extdata/demo_areas.geojson", package = "naomi"))

area_merged %>%
  filter(area_level == max(area_level)) %>%
  ggplot() +
  geom_sf(aes(geometry = geometry)) +
  theme_minimal()
```

The population data is as follows, with columns:

* `area_id`: contains the ISO, area level, and area identifier.
* `source`: where did this information come from?
* `calendar_quarter`: contains the type of year (such as CY for calendar year), the year, and the quarter
* `sex`: either female or male.
* `age_group`: in the format `Ylower_upper` where e.g. for 15-49 `lower = 15` and `upper = 49`.
* `population`: the population, which might not be an integer I guess due to survey weighting.
* `area_name`: the plain text version of the area name.
* `asfr`: age-sex fertility rate.

```{r}
pop_agesex <- read_csv(system.file("extdata/demo_population_agesex.csv", package = "naomi"))
head(pop_agesex)
```

The survey data is as follows, with columns:

* `indicator`: which survey indicator is it? Takes values `"prevalence"`, `"art_coverage"`, `"viral_suppression_plhiv"` and `"recent_infected"`.
* `survey_id`: the name of the survey, containing the country, year and type.
* `survey_mid_calendar_quarter`: which calendar quarter did the survey occur during?
* `area_id`: as before.
* `area_name`: as before.
* `res_type`: unknown.
* `sex`: as before.
* `age_group`: as before.
* `n_clusters`: the number of survey clusters in this area.
* `n_observations`: the number of individuals surveyed.
* `n_eff_kish`: the Kish effective sample size.
* `estimate`: a survey weighted estimate of the indicator.
* `std_error`: standard error of `estimate`.
* `ci_lower`: 95% lower confidence interval for `estimate`.
* `ci_upper`: 95% upper confidence interval for `estimate`.

```{r}
survey_hiv_indicators <- read_csv(system.file("extdata/demo_survey_hiv_indicators.csv", package = "naomi"))
head(survey_hiv_indicators)
```

The ART programmatic data is as follows, with columns:

* `area_id`: as before.
* `area_name`: as before.
* `sex`: as before.
* `age_group`: as before.
* `year`: the year.
* `calendar_quarter`: the calendar quarter.
* `art_current`: unknown.
* `art_new`: unknown.
* `vl_tested_12mos`: viral load tested past 12 months.
* `vl_suppressed_12mos`: viral load suppressed past 12 months.

```{r}
art_number <- read_csv(system.file("extdata/demo_art_number.csv", package = "naomi"))
head(art_number)
```

The ANC programmatic data is as follows, with columns:

```{r}
anc_testing <- read_csv(system.file("extdata/demo_anc_testing.csv", package = "naomi"))
head(anc_testing)
```

```{r}
pjnz <- system.file("extdata/demo_mwi2019.PJNZ", package = "naomi")
spec <- extract_pjnz_naomi(pjnz)
head(spec)
```

## Choose model areas and time points

* `scope`: A collection of `area_id`s defining the set of areas to be modelled. Usually this is simply national level, so the level 0 `area_id`.
* `level`: Area level at which to fit the model.
* `quarter_id_t1`: The first time point for the model -- approximately the midpoint of the household survey data used.
* `quarter_id_t2`: The second time point for the model -- the current time for which estimates are needed.
* `quarter_id_t3`: The third time point for the model -- the future projection for HIV estimates.

```{r}
scope <- "MWI"
level <- 4
calendar_quarter_t1 <- "CY2016Q1"
calendar_quarter_t2 <- "CY2018Q3"
calendar_quarter_t3 <- "CY2019Q4"
```

```{r}
prev_survey_ids  <- c("DEMO2016PHIA", "DEMO2015DHS")
artcov_survey_ids  <- "DEMO2016PHIA"
vls_survey_ids <- NULL
recent_survey_ids <- "DEMO2016PHIA"

artnum_calendar_quarter_t1 <- "CY2016Q1"
artnum_calendar_quarter_t2 <- "CY2018Q3"

anc_clients_year2 <- 2018
anc_clients_year2_num_months <- 9

anc_prevalence_year1 <- 2016
anc_prevalence_year2 <- 2018

anc_art_coverage_year1 <- 2016
anc_art_coverage_year2 <- 2018
```

## Review input data

This section is missing from the vignette!

## Prepare model inputs

Setup the model:

```{r}
naomi_mf <- naomi_model_frame(
  area_merged,
  pop_agesex,
  spec,
  scope = scope,
  level = level,
  calendar_quarter_t1,
  calendar_quarter_t2,
  calendar_quarter_t3
)
```

Prepare data inputs

```{r}
naomi_data <- select_naomi_data(
  naomi_mf,
  survey_hiv_indicators,
  anc_testing,
  art_number,
  prev_survey_ids,
  artcov_survey_ids,
  recent_survey_ids,
  vls_survey_ids,
  artnum_calendar_quarter_t1,
  artnum_calendar_quarter_t2,
  anc_prevalence_year1,
  anc_prevalence_year2,
  anc_art_coverage_year1,
  anc_art_coverage_year2
)
```

## Fit the TMB model

```{r}
tmb_inputs <- prepare_tmb_inputs(naomi_data)
fit <- fit_tmb(tmb_inputs)
```

Calculate model outputs:

```{r}
outputs <- output_package(fit, naomi_data) #' Note that in the vignette they have naomi_mf here, is this a mistake?
```

The output package consists of a data frame of indicators and metadata defining the labels for each indicator:

```{r}
names(outputs)
```

If uncertainty has not been calcualted yet, the output object retures values for the mode, but not the mean or lower and upper 95% uncertainty ranges:

```{r}
outputs$indicators %>%
  dplyr::filter(
    indicator == "prevalence",  #' HIV prevalence
    age_group == "Y015_049"   #' Age group 15-49
  ) %>%
  head()
```

The function `add_output_labels()` returns the indicators table with labels added as additional columns:

```{r}
add_output_labels(outputs) %>%
  dplyr::filter(
    indicator == "prevalence",  # HIV prevalence
    age_group == "Y015_049"   # Age group 15-49
  ) %>%
  head()
```

```{r}
system.time(fit <- sample_tmb(fit))
```

Regenerate outputs with uncertainty ranges:

```{r}
system.time(outputs <- output_package(fit, naomi_data))

outputs_calib <- calibrate_outputs(
  outputs, 
  naomi_mf,
  spectrum_plhiv_calibration_level = "national",
  spectrum_plhiv_calibration_strat = "sex_age_coarse",
  spectrum_artnum_calibration_level = "national", 
  spectrum_artnum_calibration_strat = "sex_age_coarse",
  spectrum_aware_calibration_level = "national", 
  spectrum_aware_calibration_strat = "sex_age_coarse",
  spectrum_infections_calibration_level = "national", 
  spectrum_infections_calibration_strat = "sex_age_coarse"
)

outputs$indicators %>%
  dplyr::filter(
    indicator == "prevalence",  # HIV prevalence
    age_group == "Y015_049"   # Age group 15-49
  ) %>%
  head()
```

## Plot model outputs

```{r}
indicators <- add_output_labels(outputs) %>%
  left_join(outputs$meta_area %>% select(area_level, area_id, center_x, center_y)) %>%
  sf::st_as_sf()
```

15-49 prevalence by district (which is `area_level` 4):

```{r}
indicators %>%
  filter(age_group == "Y015_049",
         indicator == "prevalence",
         area_level == 4) %>%
  ggplot(aes(fill = mode)) +
  geom_sf() +
  viridis::scale_fill_viridis(labels = scales::percent_format()) +
  th_map() +
  facet_wrap(~sex)
```

15-49 prevalence by zone (which is `area_level` 2):

```{r}
indicators %>%
  filter(age_group == "Y015_049",
         indicator == "prevalence",
         area_level == 2) %>%
  ggplot(aes(fill = mode)) +
  geom_sf() +
  viridis::scale_fill_viridis(labels = scales::percent_format()) +
  th_map() +
  facet_wrap(~sex)
```

Age-specific prevalence at a national-level:

```{r}
indicators %>%
  dplyr::filter(area_level == 0,
         sex != "both",
         age_group %in% get_five_year_age_groups(),
         calendar_quarter == "CY2018Q3",
         indicator == "prevalence") %>%
  left_join(get_age_groups()) %>%
  mutate(age_group = fct_reorder(age_group_label, age_group_sort_order)) %>%
  ggplot(aes(age_group, mean, ymin = lower, ymax = upper, fill = sex)) +
  geom_col(position = "dodge") +
  geom_linerange(position = position_dodge(0.8)) +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(labels = scales::percent_format(1)) +
  facet_wrap(~area_name) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1.0, vjust = 0.5))
```

ART coverage by age-sex-region:

```{r}
indicators %>%
  filter(area_level == 1,
         sex != "both",
         age_group %in% get_five_year_age_groups(),
         indicator == "art_coverage",
         calendar_quarter == "CY2018Q3") %>%
  left_join(get_age_groups()) %>%
  mutate(age_group = fct_reorder(age_group_label, age_group_sort_order)) %>%
  ggplot(aes(age_group, mean, ymin = lower, ymax = upper, fill = sex)) +
  geom_col(position = "dodge") +
  geom_linerange(position = position_dodge(0.8)) +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(labels = scales::percent_format(1)) +
  facet_wrap(~area_name) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1.0, vjust = 0.5))
```

Bubble plot of prevalence and PLHIV:

```{r}
indicators %>%
  filter(age_group == "Y015_064",
         area_level == 4,
         indicator %in% c("prevalence", "plhiv"),
         calendar_quarter == "CY2018Q3") %>%
  select(sex, center_x, center_y, indicator_label, mean) %>%
  spread(indicator_label, mean) %>%
  ggplot() +
  geom_sf() +
  geom_point(aes(center_x, center_y, colour = `HIV prevalence`, size = PLHIV)) +
  viridis::scale_color_viridis(labels = scales::percent_format()) +
  th_map() +
  facet_wrap(~sex)
```
