---
title: "INLA two dimensional grid replication"
author:
- name: Adam Howes
output:
  html_document:
    toc: yes
    toc_float: true
    toc_collapsed: true
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
    **Background** The INLA method uses a grid of hyperparameters to integrate over.
    
    **Task** We replicate a method for selecting grid vaues as outlined in Rue, Martino and Chopin (2009).
---

## A two dimensional MVN

Suppose that $\theta = (\theta_1, \theta_2)$ has a multivariate Gaussian distribution with mean vector $\mu$ and covariance matrix $\Sigma$ as follows:

```{r}
mu <- c(1, 1.5)
cov <- matrix(c(2, 1, 1, 1), ncol = 2)
```

The probability density function (PDF) is:

```{r}
obj <- function(theta) {
  mvtnorm::dmvnorm(theta, mean = mu, sigma = cov)
}

grid <- expand.grid(
  theta1 = seq(-2, 5, length.out = 700),
  theta2 = seq(-2, 5, length.out = 700)
)

ground_truth <- cbind(grid, pdf = obj(grid))

ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) +
  geom_contour(col = multi.utils::cbpalette()[1]) +
  coord_fixed(xlim = c(-2, 5), ylim = c(-2, 5), ratio = 1) +
  theme_minimal()
```

Find the mode via optimisation (the answer is clearly `mu` but to be complete):

```{r}
opt <- optim(par = c(0, 0), fn = obj, control = list(fnscale = -1))
opt$par
```

The Hessian is given by `cov` (should compute this numerically).
Using the Eigendecomposition $\Sigma = V \Lambda V^\top$:

```{r}
Lambda <- diag(eigen(cov)$values) #' Diagonal matrix
V <- eigen(cov)$vectors #' Eigenvector matrix
V %*% Lambda %*% t(V) #' Verifying that this is cov
```

## Explore the $z$-coordinates

Create a mapping $\theta(z)$ from the $z$-coordinates to $\theta$-coordinates:

```{r}
z_to_theta <- function(z) {
  as.vector(opt$par + V %*% sqrt(Lambda) %*% z)
}
```

The origin `c(0, 0)` maps to the mode:

```{r}
theta_mode <- z_to_theta(c(0, 0))
theta_mode

test_statistic <- function(theta_proposal) {
  log(obj(theta_mode)) - log(obj(theta_proposal))
}
```

Create a function to explore in $z$-coordinates along one dimension:

```{r}
#' @param m is dim(theta)
#' @param j is the index of exploration direction
#' @param delta_z is the step size
#' @param delta_pi is the acceptable drop-off
explore_direction <- function(j, m, delta_z, delta_pi) {
  z_mode <- rep(0, m)
  z_names <- paste0("z", 1:m)
  names(z_mode) <- z_names

  unit_vector <- rep(0, m)
  unit_vector[j] <- 1

  points <- rbind(z_mode)

  # Increasing
  i <- 0
  condition <- TRUE
  while(condition) {
    i <- i + 1
    proposal <- c(z_mode + i * delta_z %*% unit_vector)
    names(proposal) <- z_names
    statistic <- test_statistic(z_to_theta(proposal))
    condition <- (statistic < delta_pi)
    if(condition){
      points <- rbind(points, proposal)
    }
  }

  # Decreasing
  i <- 0
  condition <- TRUE
  while(condition) {
    i <- i + 1
    proposal <- c(z_mode - i * delta_z %*% unit_vector)
    names(proposal) <- z_names
    statistic <- test_statistic(z_to_theta(proposal))
    condition <- (statistic < delta_pi)
    if(condition){
      points <- rbind(points, proposal)
    }
  }
  as.data.frame(points)
}
```

I chose values for the step size and drop-off manually based on what looked good:

```{r}
d_z <- 0.2
d_pi <- 2
```

Expand the scaffold in $z$-coordinates:

```{r}
z_grid <- expand.grid(
  z1 = explore_direction(1, 2, delta_z = d_z, delta_pi = d_pi)$z1,
  z2 = explore_direction(2, 2, delta_z = d_z, delta_pi = d_pi)$z2
)
```

Map it back to $\theta$-coordinates:

```{r}
theta_grid_full <- t(apply(z_grid, 1, z_to_theta)) %>%
  as.data.frame() %>%
  rename(theta1 = V1, theta2 = V2)
```

Keep only those points which meet condition:

```{r}
theta_grid_full %>%
  mutate(statistic = apply(theta_grid_full, 1, test_statistic),
         condition = statistic < d_pi) %>%
  filter(condition == TRUE) -> theta_grid
```

```{r}
#' Eigenvector line segments
rbind(c(theta_mode[1], theta_mode[2], (theta_mode + V[,1])[1], (theta_mode + V[,1])[2], 1),
      c(theta_mode[1], theta_mode[2], (theta_mode + V[,2])[1], (theta_mode + V[,2])[2], 2)) %>%
  as.data.frame() %>%
  rename(x = V1, y = V2, xend = V3, yend = V4, pc = V5) -> segments

plot <- ggplot() +
  geom_point(data = theta_grid, aes(x = theta1, y = theta2), alpha = 0.5) +
  geom_contour(data = ground_truth, aes(x = theta1, y = theta2, z = pdf), col = multi.utils::cbpalette()[1]) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend, colour = as.factor(pc)),
               data = segments, size = 1, arrow = arrow(length = unit(0.3, "inches")),
               lineend = "round", linejoin = "round") +
  scale_colour_manual(values = multi.utils::cbpalette()[-1]) +
  labs(x = "theta1", y = "theta2") +
  theme_minimal() +
  theme(legend.position = "none")

plot
```

```{r}
pdf("inla-grid.pdf", h = 4, w = 6.25)

plot

dev.off()
```
